---
author: Miguel Camacho Sánchez
title: "tidyGenR benchmarking"
editor: visual
theme: minty
format:
  html:
    toc: true
    code-tools: true
---

Attach libraries:

```{r}
#| echo: false
#| include: false
#knitr::opts_knit$set(root.dir = "../")
```

```{r attach packages}
#| message: false
#| warning: false
library(tidyGenR)
library(dplyr)
library(patchwork)
library(ggplot2)
```

```{r eval_conditionals}
#| echo: false
#| eval: true
# declare logicals to trigger EVAL in intensive code chunks
# they are not run if output already exists.
# to run a fresh analysis outputs need to be removed.
# demultiplex
c_log <- "output/cutadapt.log"
eval_demultiplex <- !file.exists(c_log)
# fastqc
multiqc_report <- "output/multiqc_report.html"
eval_fastqc <- !file.exists(multiqc_report)
# truncate
tr_path <- "output/trunc_in-out.rds"
eval_truncate <- !file.exists(tr_path)
# explore_dada
x_out_path <- "output/explore_dada.rds"
eval_x <- !file.exists(x_out_path)
# variant_call
x_variants_path <-  "output/variants_x.rds"
eval_variant_call <- !file.exists(x_variants_path)
# genotype
x_genotypes_path <-  "output/genotypes_x.rds"
eval_genotype <- !file.exists(x_genotypes_path)
# amplisas
amplisas_results <- ""
eval_amplisas <- !file.exists(amplisas_results)
```

## Benchmarking *tidyGenR*

This repository accompanies the R package **tidyGenR** available at <https://github.com/csmiguel/tidyGenR>. It covers (1) all steps for genotype calling with *tidyGenR*, (2) exploration of the parameters for variant calling, and (3) comparison of *tidyGenR* against [AmpliSAS](https://doi.org/10.1111/1755-0998.12453). It uses real data from a population genetics study of the wild rodent *Rattus baluensis* (Camacho-Sanchez et al. *in preparation*).

## Preparation of input data

Raw sequences are deposited in NCBI under the BioStudy [SRP293699](https://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP293699), in the BioProject [PRJNA680166](https://www.ncbi.nlm.nih.gov/bioproject/PRJNA680166). From the BioStudy, we can download the [SRA metadata](data/raw/SraRunTable.csv) to map sample IDs to SRRs:

```{r SRR}
# table with mapped SRR-sampleIDs
srr <-
  read.csv("data/raw/SraRunTable.csv") |>
  select(Run, "Sample.Name") |>
  rename(run = Run,
         sample = "Sample.Name")
knitr::kable(head(srr, 3))
```

The FASTQ reads from the SRRs can be downloaded with [SRA-toolkit](https://github.com/ncbi/sra-tools):

```{r download_raw_data}
#|  message: false
#|  warning: false
#|  error: false
#|  code-fold: true
# download raw reads from NCBI
# they are not downloaded if already present in 'data/raw'.
apply(srr, 1, function(x) {
  # name paths
  srr_path <- file.path("data/raw", x[1])
  srr1 <- file.path("data/raw", paste0(x[1], "_1.fastq"))
  srr2 <- file.path("data/raw", paste0(x[1], "_2.fastq"))
  sam1 <- file.path("data/raw", paste0(x[2], "_1.fastq"))
  sam2 <- file.path("data/raw", paste0(x[2], "_2.fastq"))
  # fetch SRR
  if (!any(dir.exists(srr_path) || file.exists(sam1))) {
    system2("prefetch", paste("-O", "data/raw/", x[1]))
  }
  # reformat to FASTQ
  if (!file.exists(sam1)) {
    system2("fasterq-dump", paste("-O", "data/raw/", srr_path))
    system2("mv", paste(srr1, sam1))
    system2("mv", paste(srr2, sam2))
    # rm SRR
    system2("rm", paste0("-r ", srr_path, "*"))
  }
  })
# glimpse raw FASTQ
list.files("data/raw", pattern = "fastq")[1:3]
```

Check input data files:

```{r check_raw_reads}
freads <- list.files("data/raw", pattern = "1.fastq",
                    full.names = TRUE)
rreads <- list.files("data/raw", pattern = "2.fastq",
                    full.names = TRUE)
chr <- check_raw_reads(freads, rreads, low_readcount = 10)
```

Input FASTQ is conditions for *tidyGenR*. A total of `r length(chr$samples)` are detected:

```{r}
chr$samples
```

## Demultiplex by locus

Reads are demultiplexed by locus using primer sequences in paired-end mode.

```{r demultiplex_objects}
#| warning: false
# load primer data
data("primers")
# path to cutadapt
cutadapt <- system("which cutadapt", intern = TRUE)
# path to folder save locus-demultiplexed FASTQ
demult <- "data/intermediate/demultiplexed"
# print primers
knitr::kable(head(primers, 3))
```

```{r demultiplex}
#| eval: !expr eval_demultiplex
#| warning: false
# demultiplex
demultiplex(
  interpreter = "/bin/bash",
  cutadapt = cutadapt,
  freads = freads,
  rreads = rreads,
  primers = primers,
  sh_out = "code/demultiplex.sh",
  write_demultiplexed = demult,
  log_out = c_log,
  mode = "pe",
  run = TRUE)
```

```{r}
# glimpse demultiplexed FASTQ
list.files(demult, pattern = "fastq")[1:3]
```

Remove files with few reads:

```{r rm_after_dem}
#| warning: false
remove_poor_fastq(demult,
                   min_reads = 10)
```

Make sequence-quality reports using FastQC and MultiQC. Instead of running it over the >2000 demultiplexed files, it will be run over 100 random files.
```{r}
#| eval: !expr eval_fastqc
# instead of running fastqc to all files > 2000, I run
#    it on 100 random files.
dem_files <-
  list.files(demult, "fastq.gz", full.names = TRUE)
# select 100 random files
random_dem <-
  sample(dem_files, 100, replace = FALSE)
# copy to temp dir
temp_dem <- file.path(tempdir(), "dem")
dir.create(temp_dem)
file.copy(from = random_dem, to = temp_dem)
# run fastqc
system2("fastqc",
        paste("--noextract -o",
              temp_dem,
              paste0(temp_dem, "/*fastq.gz")))

# run multiqc
system2("multiqc",
        paste("-o output", temp_dem))
           
```

## Truncate reads

Reads are truncated to a given length for each locus. The truncation lengths depend on the sequence qualities for forward and reverse reads. After per-base sequence-qualities in the `output/multiqc_report.html`, **270** nt for forward reads and **180** for reverse reads seem reasonable truncation lengths. For some loci, specific truncation lengths were set to maximize the number of reads returned. That is, a `data.frame` with truncation lengths for forward and reverse reads was built to maximize the amount of information yielded by each locus. For instance, the amplicons for some loci, as *nfkbia*, are long and it is a good trade-off to leave the low quality ends in but to be sure both F and R reads overlap.

Building of `data.frame` with locus-specific truncation lengths:

```{r build_trunc_fr}
#| message: false
#| include: false
loci <- 
  tidyGenR:::check_names_demultiplexed(demult,
                                     fw_pattern = "1.fastq.gz",
                                     rv_pattern = "2.fastq.gz")$loci

trunc_fr <-
  data.frame(locus = loci,
             trunc_f = 270,
             trunc_r = 180)
# introduce manual values
trunc_fr[which(loci == "rgd735029"), "trunc_f"] <-  245
trunc_fr[which(loci == "rgd735029"), "trunc_r"] <-  155
trunc_fr[which(loci == "fancg"), "trunc_r"] <- 190
trunc_fr[which(loci == "nfkbia"), "trunc_r"] <- 240
trunc_fr[which(loci == "tmem87a"), "trunc_r"] <- 160

# glimpse data.frame with locus-specific truncation lengths
knitr::kable(head(trunc_fr, 3))
```

Truncate reads according to locus-specific truncation lengths:

```{r declare_tr_dir}
tr_dir <- "data/intermediate/truncated"
```

```{r truncate}
#| message: false
#| warning: false
#| eval: !expr eval_truncate

# truncate
trunc_out <-
  trunc_amp(in_dir = demult,
          fw_pattern = "1.fastq.gz",
          rv_pattern = "2.fastq.gz",
          trunc_fr = trunc_fr,
          write_trun = tr_dir,
          max_ee = c(4, 5),
          trunc_q = 2)
# save reads in and out
saveRDS(trunc_out, tr_path)
```

The output from `trunc_amp()` is a list of matrices with IN and OUT reads after truncation:

```{r}
#| echo: false
trunc_out <- readRDS(tr_path)
remove_poor_fastq(tr_dir,
                   min_reads = 10)
```

```{r}
# see trunc_out
lapply(trunc_out[seq_len(3)], head, 3)
```

## Exploration of the parameter space

The function `explore_dada()` can be used to explore the effect of DADA2 parameters ("OMEGA_A", "BAND_SIZE", "pool") on the sensitivity on the variant calling.

-   *omega_a*: threshold for variants to be significant overabundant *log(-log(birth_pval))* (see [Rosen et al. 2012](https://doi.org/10.1186/1471-2105-13-283.)).
-   *band_size*: positive numbers set a band size in Needleman-Wunsch alignments. In this context, ends free alignment is performed. A value of zero turns off banding, triggering full Needleman-Wunsch alignments, in which gapless alignment is performed (see [issue](https://github.com/benjjneb/dada2/issues/1982)).
-   *pool*: calling variants pooling samples can increase sensitivity (see [dicussion](https://benjjneb.github.io/dada2/pseudo.html)).

The returned plots can be used to guide the election of the best *OMEGA_A* in `variant_call()`, or frequency (*maf*) and abundance thresholds (*ad*) for filtering variants.

Explore variants:

```{r}
# declare candidate OMEGA_A to use in variant_call()
candidate_omega_a <- 10^-2
```

```{r explore_dada}
#| eval: !expr eval_x
#| message: false
# paths to forward and reverse truncated reads
ftrun <-
  list.files(tr_dir, pattern = "_F_", full.names = T)
rtrun <-
  list.files(tr_dir, pattern = "_R_", full.names = T)
# candidate omega value to annotate vline in plots.
v_line <- log(-log(candidate_omega_a))
# run explore_dada() with band_size = 0, non pooling and omega_a = 0.9
# forward
F_ind_0 <-
  explore_dada(ftrun, band_size = 0, vline = v_line, hline_fr = 0.1)
# reverse
R_ind_0 <-
  explore_dada(rtrun, band_size = 0, vline = v_line, hline_fr = 0.1)
# save results
saveRDS(list(F_ind_0 = F_ind_0, R_ind_0 = R_ind_0), x_out_path)
```

Exploration of DADA2 clustering for forward (A, C) and reverse (B, D) reads. The Y-axis represents the frequency of the variant in each locus and sample. The *log(-log(birth_pval))* transformation in the X-axis is related to the p-value of a variant being significantly overabundant. Larger x-values represent likely true variants. For representation purposes *birth_pval* of 0 (thus negative infinite), are converted to 10. Points are color-coded according to the variant rank in read abundance for its given locus and sample. For diploid individuals, [green]{style="color: green;"} are likely true variants and [red]{style="color: red;"} are likely false variants. [Grey]{style="color: grey;"} dashed lines are thresholds used for `variant_call()`:

```{r x_attach_if_not_run}
#| echo: false
#| eval: true
if (!(exists("F_ind_0") || exists("R_ind_0"))) {
    x_dada <- readRDS(x_out_path)
    attach(x_dada)
}
```

```{r plot_explore_dada}
#| label: fig-explore_dada
#| warnings: false
#| fig-cap: "Exploration of DADA2 variants"
#| code-fold: true
ppool <-
  (F_ind_0$p1 | R_ind_0$p1) / (F_ind_0$p2 | R_ind_0$p2) +
  patchwork::plot_annotation(title = "Dada F/R pool = F, omega_a 0.9, band_size = 0",
                             tag_levels = "A")
# save plot with combined loci
ggsave("output/explore_dada.pdf", ppool, width = 8, height = 5)
# print plot
ppool
```

After the exploration variants in @fig-explore_dada, it seems an *OMEGA_A* = `r candidate_omega_a`, implying a cutoff of `r log(-log(candidate_omega_a))` in the X-axis and a frequency threshold (Y-axis) of 0.1, excludes most [artifacts]{style="color: red;"} while maximizing [true positives]{style="color: green;"}.

The results can also be explored **per-locus**. For instance, @fig-explore_dada B can be expanded per locus as follows:

```{r}
#| include: false
#| message: false
#| code-fold: true
# list of plots per locus
lplots <-
  list(loci_f_ind_0_logp = F_ind_0$p3,
       loci_r_ind_0_logp = R_ind_0$p3,
       loci_f_ind_0_abun = F_ind_0$p4,
       loci_r_ind_0_abun = R_ind_0$p4)

# save plots per locus
lapply(seq_along(lplots), function(x) {
  ggsave(paste0("output/", names(lplots)[x], ".pdf"),
         lplots[[x]], width = 6, height = 6)
  })
```

```{r}
#| label: fig-explore_dada_loci
#| warnings: false
#| fig-cap: "Exploration of DADA2 variants per locus for R reads"
lplots$loci_r_ind_0_logp
```

## Variant and genotype calling

Variant calling is run using *OMEGA_A* = `r candidate_omega_a`, and under different parameters:

-   *band_size*: 0, 16
-   *pool*: TRUE, FALSE

```{r variant_call}
#| message: false
#| eval: !expr eval_variant_call
#| include: false
# list of parameters to run variant_call() with 4 different set of values:
variant_call_params <-
  list(
    ind_bs16 = c(FALSE, 16), # pool = F, default band size == 16
    ind_bs0 = c(FALSE, 0), # pool = F, default band size == 0
    pool_bs16 = c(TRUE, 16), # pool = T, default band size == 16
    pool_bs0 = c(TRUE, 0) # pool = T, default band size == 0
)

variants_x <-
  lapply(variant_call_params, function(x) {
    variant_call(in_folder = tr_dir,
                 rv_pattern = "R_filt.fastq.gz",
                 c_unmerged = TRUE,
                 multithread = TRUE,
                 pool = as.logical(x[1]),
                 omega_a_f = candidate_omega_a,
                 omega_a_r = candidate_omega_a,
                 band_size = x[2],
                 ad = 10,
                 maf = 0.1)
  })
saveRDS(variants_x, "output/variants_x.rds")
```

```{r}
#| echo: false
if (!exists("variants_x"))
  variants_x <- readRDS("output/variants_x.rds")
```

Variants are used to genotype individuals with defaults `ploidy = 2` and `ADt = 10`:

```{r}
#| message: false
#| eval: !expr eval_genotype
genotypes_x <-
  lapply(variants_x, genotype)
saveRDS(genotypes_x, file = "output/genotypes_x.rds")
```

```{r}
#| echo: false
if (!exists("genotypes_x"))
  genotypes_x <- readRDS("output/genotypes_x.rds")
```

## Tidy data

A strong point from *tidyGenR* is that variants and genotypes from `variant_call()` and `genotype()` are returned in `tidy` format: one row per observation and variables in columns. Lets have a look at the data structure.

```{r glimpse_tidy_variants}
# glimpse tidy variants
knitr::kable(head(variants_x$ind_bs0))
```

```{r glimpse_tidy_genotypes}
# glimpse tidy genotypes
knitr::kable(head(genotypes_x$ind_bs0))
```

## Genotype with AmpliSAT

For later comparison with *tidyGenR*, genotyping is carried with [AmpliSAT](https://doi.org/10.1111/1755-0998.12453), a software written in PERL with similar characteristics to *tidyGenR*. To compare their performance we run AmpliSAT in a docker container to genotype our raw sequences. The steps are detailed [here](code/amplisas). AmpliSAS returns results in a multisheet EXCEL and in plain text files, one per locus. The function `amplisas2tidy()` permits to read plain text results from AmpliSAS into tidy variants.

Create input for AmpliSAS:

```{r}
#| eval: false
# amplicon metadata
source("code/amplisas/01_create_amplicon_data.R")
# append barcodes to reads
source("code/amplisas/02_add_barcodes.R")

```

AmpliSAS is run in a DOCKER container. The steps are detailed [here](code/amplisas/amplisas_analysis.Rmd).

AmpliSAS results can be read to tidy variants:

```{r read_amplisas}
#| message: false
fp <- list.files("data/intermediate/amplisas/results_amplisas/filtered",
                 pattern = "txt$",
                 full.names = TRUE)
# amplisas restuls to tidy variants
var_amplisas <-
  amplisas2tidy(fp)

# to genotypes
gen_amplisas <-
  genotype(var_amplisas)
knitr::kable(head(var_amplisas, 3))
```

## Compare results

*compare_calls()* is able compare tidy variants and tidy genotypes data frames. It generates tables and plots describing the differences between objects. It is useful to compare results after running *tidyGenR* with different parameter sets. In this case, I will carry out two comparisons:
  + 1. genotypes: 4 runs of tidyGenR plus AmpliSAT.
  + 2. genotypes: AmpliSAT plus the most similar run.

The comparison of the four *tidyGenR* and AmpliSAT indicate that set of parameters that returns the most similar genotype calls to AmpliSAT is using a band size = 0.

```{r}
# compare genotypes of the four runs plus amplisat
#   genotypes
gen_comb <-
  c(genotypes_x,
    list(gen_amplisas))
names(gen_comb) <- c(stringr::str_c("tidyGenR_", names(genotypes_x)), "AmpliSAT")
comp_gen <- compare_calls(gen_comb,
                "output/comp_gen_noReads.xlsx",
                creads = FALSE)
saveRDS(comp_gen, "output/comp_gen.rds")
knitr::kable(comp_gen$dist)
```

A plot comparing the genotypes called was used to investigate differences between strategies and explore potential biases and problematic loci.

```{r}
#| fig-cap: "Comparison between genotype calls."
#| label: "comp_gen_all"
#| code-fold: true
comp_gen$plot2
```
Problematic loci, with low coverage and with conflicting genotypes between different strategies were checked manually by aligning them and compared the with [dereplicated sequences](output/dereplicated.xlsx). Setting band_size = 0 yielded the best results as it recovered alleles which had indels at their ends. Pooling samples did not affect genotype calls, although it has some influence on the number of reads supporting each allele (turn on `creads` to explore this case). This confirms that samples that do not share allele (i.e. different species) can be genotyped equally efficiently as samples that share allele (i.e. same species). So, the run with band size = 0 and non-pooled samples was chosen as the best strategy and compared pairwise with AmpliSAT:

```{r}
# compare best call with amplisas
#   variants
comp_gen_sel <-
  compare_calls(list(tidyGenR_ind_bs0 = genotypes_x$ind_bs0,
                      AmpliSAT = gen_amplisas),
                  "output/comp_gen_indbs0_amplisat.xlsx",
                  creads = TRUE)
saveRDS(comp_gen_sel, "output/comp_gen_sel.rds")

```


```{r plot_amplisat_vs_tidygenr}
#| label: fig-compgen-amplisas
#| warnings: false
#| fig-cap: "Comparison of genotype call tidyGenR vs AmpliSAT"
#| code-fold: true
pcomp1 <- comp_gen_sel$plot2
ggsave(pcomp1,
       file = "output/comp_amplisat_indo.pdf",
       width = 10,
       height = 5.4)
pcomp1
```

```{r plot_n_reads_comp}
#| label: fig-compgen-amplisas_nreads
#| warnings: false
#| fig-cap: "Comparison of number of reads supporting alleles in tidyGenR vs AmpliSAT"
#| code-fold: true
pcomp2 <- comp_gen_sel$plot3
ggsave(pcomp2,
       file = "output/comp_amplisat_ind0_boxplot.pdf",
       width = 5,
       height = 3)
pcomp2
```
Lastly, the genotypes can be compared against their respective variants to have an idea of variants being dropped when genotyping:
```{r drop_variants}
#| label: fig-dropped_var
#| warnings: false
#| fig-cap: "Identified variants dropped during genotyping"
#| code-fold: true
dropped_var <-
      list(variants = variants_x$ind_bs0,
            genotypes = genotypes_x$ind_bs0)
comp_var_gen <- compare_calls(dropped_var)
saveRDS(comp_var_gen, "output/comp_var_gen.rds")
comp_var_gen$plot1

```

# session info

Details on the sessionInfo() are [here](session_info.txt):

```{r}
writeLines(capture.output(sessionInfo()), con = "session_info.txt")
```
